---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rairtable

<!-- badges: start -->
[![R-CMD-check](https://github.com/matthewjrogers/rairtable/workflows/R-CMD-check/badge.svg)](https://github.com/matthewjrogers/rairtable/actions)
[![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/rairtable)](https://cran.r-project.org/package=rairtable)
[![metacran downloads](https://cranlogs.r-pkg.org/badges/rairtable)](https://cran.r-project.org/package=rairtable)
<!-- badges: end -->

`rairtable` is an efficient interface to the Airtable API intended to simplify the integration of Airtable into data science workflows. Using `rairtable` you can:

- Set and get Airtable API keys and personal access tokens
- Create, update, and delete Airtable records
- Use Airtable as an endpoint for `dplyr` pipelines with tidyverse-friendly functions
- Get filtered data from Airtable views

## Installation

For the stable CRAN release, you can install rairtable with:

``` r
install.packages('rairtable')
```

You can install the development version of rairtable from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pkg_install("matthewjrogers/rairtable@dev")
```

## Usage

```{r}
library(rairtable)
```

### Get and set a personal access token (PAT)

Create a personal access token [using your Airtable account](https://airtable.com/create/tokens) to access the API.

Save the token to your local environment using `set_airtable_api_key()` and set `install = TRUE` to save the token for reuse in future sessions:

``` r
set_airtable_pat('<your key goes here>')

set_airtable_pat('<your key goes here>', install = TRUE)
```

This package does support Airtable API keys, however, Airtable support for API keys [ends in January 2024](https://support.airtable.com/docs/airtable-api-key-deprecation-notice). Any script using an API key should be updated to use a personal access token instead.

### Create an `airtable` object and reading records

This README uses a sample Airtable copied from the [Emojis Database](https://www.airtable.com/universe/exphjm5ifnV0bX4Kb/emojis-database) in the [Airtable Universe](https://www.airtable.com/universe). 

Use an Airtable base and table ID to create an `airtable` object:

```{r}
atbl <- airtable(
  base = "app8uSq5M3ia6FZUR",
  table = "tblvGyhX2yg0d22gz"
)

atbl
```

The airtable object is a list that holds your base ID and a request URL. 

You can also use the URL for your base to create an airtable object. Note, that the URL typically includes a view as well as a table. If a view is filtered to only show a subset of records, only the records you can see in that view can be listed if you pass the airtable object to `read_records()` or `list_records()`.

```{r}
atbl <- airtable(
  table = "https://airtable.com/app8uSq5M3ia6FZUR/tblvGyhX2yg0d22gz/viw4HALSnD2HQMOHs"
)
```

You can read records to a data frame using `read_airtable()`:

```{r}
records <- read_airtable(airtable = atbl)

records
```

By default, `read_airtable()` will read all rows in the chosen table and store Airtable records IDs in a column named `airtable_record_id`. Optionally, airtable record IDs can be stored as row names for convenience. However, this is not recommend in most cases since some operations such as `dplyr::arrange()` will drop row names. You can record ID column name by setting the option for `rairtable.id_col`.

You can get a single record with `get_record()`:

```{r}
get_record(atbl, record = "rec0GH7OYgPEjRaSb")
```

`list_records()` is similar to `read_airtable()` but supports some additional features from the API.

### Insert records 

Records can be added to an Airtable base by passing a data frame to `insert_records()` or `create_records()` (the two functions are identical):

```{r}
data <- tibble::tibble(
  "Emoji" = "🤖",
  "Description" = "README robot"
)

insert_records(
  airtable = atbl,
  data = data
)
```

The functions do not currently validate data so the column names and types must match the names and types in the Airtable where the records will be created. You can also delete records using `delete_records()`.

### Access the Airtable Metadata API

The Airtable Metadata API requires a personal access token and will not work with an API key. The `airtable_base()` function returns a list containing the base ID, `airtable_base_schema` object, and an airtable object for each table in the base.

```{r}
airtable_base(atbl)
```

`list_base_tables()` is more basic alternative by simply returning a tibble of tables with list columns for views and fields associated with each table.

```{r}
list_base_tables(atbl)
```

